Descriptive, not injunctive, social norms caused increases in mask wearing during the COVID-19 pandemic

### 2022-12-19 manuscript.Rmd
submission

---
title             : "Descriptive, not injunctive, social norms caused increases in mask wearing throughout the COVID-19 pandemic"
shorttitle        : "Norms and mask wearing"

author: 
  - name          : "Samantha L. Heiman"
    affiliation   : "1"
    equal_contrib : yes
    corresponding : yes
    address       : "1101 E 10th St, Bloomington, IN 47405, United States"
    email         : "slheiman@iu.edu"
  - name          : "Scott Claessens"
    affiliation   : "2"
    equal_contrib : yes
  - name          : "Jessica D. Ayers"
    affiliation   : "3"
  - name          : "Diego Guevara Beltran"
    affiliation   : "3"
  - name          : "Andrew Van Horn"
    affiliation   : "4"
  - name          : "Edward R. Hirt"
    affiliation   : "1"
  - name          : "Athena Aktipis"
    affiliation   : "3"
  - name          : "Peter M. Todd"
    affiliation   : "1,5"

affiliation:
  - id            : "1"
    institution   : "Department of Psychological and Brain Sciences, Indiana University Bloomington, United States"
  - id            : "2"
    institution   : "School of Psychology, University of Auckland, New Zealand"
  - id            : "3"
    institution   : "Department of Psychology, Arizona State University, United States"
  - id            : "4"
    institution   : "Department of Anthropology, Rutgers University - New Brunswick, United States"
  - id            : "5"
    institution   : "Cognitive Science Program, Indiana University Bloomington, United States"

abstract: |
 Social norms allow humans to coordinate and cooperate in the face of existential threats. In particular, injunctive social norms prescribe what people _ought_ to do, whereas descriptive social norms inform what people _actually_ do. While previous experimental work has revealed people’s sensitivity to normative influence, several open questions remain about the natural emergence of injunctive and descriptive social norms within populations and their causal influences on cooperative behavior over time. To understand how social norms emerge and shape behavior in a non-experimental setting, we studied mask wearing during the COVID-19 pandemic. Leveraging two years of longitudinal data from a representative sample of adults in the United States (18 time points; _n_ = `r nrow(tar_read(d))`), we tracked people’s reported mask wearing behavior and their perceived injunctive and descriptive mask wearing norms as the pandemic unfolded. Longitudinal trends of norm perceptions and self-reported mask wearing suggested that norms and behavior were tightly coupled and both changed quickly in response to recommendations from public health authorities. In addition, a random-intercept cross-lagged panel model revealed that descriptive norms, but not injunctive norms, caused future increases in mask wearing behavior. These findings underline the relative importance of descriptive social norms in shaping cooperative behavior during uncertain times.

note              : "This working paper has not yet been peer-reviewed. This study was funded by the Interdisciplinary Cooperation Initiative, ASU President's Office, the Cooperation Science Network, the Institute for Mental Health Research, the University of New Mexico, the Indiana University College of Arts & Sciences, the Rutgers University Center for Human Evolutionary Studies, the Charles Koch Foundation, and the John Templeton Foundation."
keywords          : "descriptive norms; injunctive norms; longitudinal; COVID-19; mask wearing; cooperation"
wordcount         : "4683 words"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_word
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
appendix          : "appendix.Rmd"
csl               : pnas.csl
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
tar_load_everything()
# to ensure that the Rmd runs last, load appendix files again
tar_load(c(d, plot1, plot2, plot4, plot5, plot6,
           m1.1, m1.2, itemTable, lavaanTable,
           changePointsTable, constructValDes, constructValInj))
```

Social norms are a key aspect of human sociality. Broadly defined as commonly known behavioral guidelines enforced by groups of people, social norms have attracted the attention of behavioral scientists for decades [@Bicchieri2009; @Cialdini1991; @Fehr2018a; @Legros2020; @VanKleef2019]. Since Asch's early studies of normative conformity [@Asch1956], evidence has accrued that humans are particularly attuned to social norms [@Tomasello2014]. From a young age, children begin to adhere to and enforce group-wide social norms [@Jensen2014; @Rakoczy2013; @Schmidt2012] and both children and adults rely on normative emotions, such as shame and guilt, to determine when they or others have violated social norms [@Vaish2011; @Schaumberg2022]. This uniquely human sensitivity to social norms allows groups of unrelated people to cooperate and coordinate in the face of existential threats, such as resource scarcity, natural disasters, and infectious diseases [@Gelfand2011; @Henrich2015].

Previous research has distinguished between two primary kinds of social norms: injunctive norms and descriptive norms [@Bicchieri2009; @Cialdini1991; @Deutsch1955]. Injunctive social norms indicate what others in the group tend to approve or disapprove of and often involve social sanctions if violated. By contrast, descriptive norms are situational, simply describing what most people are doing in a given situation. Though these two kinds of social norms tend to align, they can also be in conflict with one another. For example, there may be an injunctive norm that cleaning up litter at a picnic site is the right thing to do: one _ought_ to behave this way. However, if an individual observes that most people are leaving their litter behind at the site, the descriptive norm is to not clean up. It is thus possible for injunctive and descriptive norms to have independent effects on behavior [@Schultz2007].

Despite decades of research on the causes and consequences of injunctive and descriptive norms [@Bicchieri2009; @Cialdini1991; @Cialdini2007; @Cialdini2021; @Schultz2007], social norms have remained an elusive concept in the behavioral sciences and several open questions remain [@Fehr2018a; @Legros2020; @VanKleef2019]. In the current work, we focus on two such questions. First, how do injunctive and descriptive norms emerge naturally over time within a population? Second, how do evolving injunctive and descriptive norms causally influence behavior over time?

With regards to norm emergence, research in cultural evolution and behavioral economics have begun to illuminate how social norms emerge over time. In the longer term, cultural evolutionary models show that injunctive social norms can be vertically transmitted down through generations via imitation or teaching, or horizontally diffused from neighboring human populations [@Boyd1985; @Henrich2015]. For example, cultural phylogenetic studies have revealed patterns of vertical cultural inheritance across societies for a variety of injunctive social norms, such as norms governing land ownership [@Kushnick2014] and post-marital residence [@Jordan2009]. However, much less is known about how social norms arise endogenously within populations in the shorter term. Recent experimental work in behavioral economics suggests that social norms of public good provisioning develop in tandem with cooperative behavior through repeated interactions [@Titlestad2019] and require peer enforcement to become stable [@Fehr2018b]. But it remains unclear to what extent these findings generalize beyond the laboratory to real human populations.

With regards to normative influences on behavior, there is a wealth of cross-sectional evidence demonstrating the behavioral impact of social norms. For example, field experiments have demonstrated the positive effects of descriptive norms on a variety of cooperative behaviors, including recycling [@Nigbur2010], paying taxes [@Larkin2018], and sustainably reusing towels in hotels [@Goldstein2008]. Evidence also suggests that any potentially deleterious effects of descriptive social norms (e.g., choosing to litter at a picnic site that already contains visible signs of littering) can be counteracted by instead focusing individuals’ attention on injunctive norms [@Schultz2007]. However, these cross-sectional studies have two main limitations. First, studies have not adequately controlled for other potential non-social influences on behavior, such as factual beliefs and personal normative beliefs [@Bicchieri2014]. For example, social norms might only affect cooperative behavior indirectly by influencing perceptions of the effectiveness of the behavior (i.e., factual beliefs) or individuals' moral preferences (i.e., personal normative beliefs). Second, cross-sectional studies have tended to follow experimental designs in which perceptions of social norms are manipulated by the researchers at a single time point, and thus do not allow cooperative social norms to emerge and affect behavior naturally over time within a population. An alternative way to identify causality, whilst retaining ecological validity, is to follow perceptions of social norms and cooperative behavior over time amidst a real, unfolding social dilemma. This is particularly important because social norms are not static: they change dynamically over time through processes of deliberation and social interaction [@Smith2015].

To understand how novel injunctive and descriptive social norms emerge over time and shape cooperative behavior in a non-experimental setting, we studied mask wearing behavior during the COVID-19 pandemic. Before the pandemic, mask wearing was not a common behavior in the United States. In April 2020, two months into the pandemic, mask wearing was officially recommended by the Centers for Disease Control and Prevention (CDC) as a cooperative behavior that people should adopt to minimize the spread of COVID-19 [@CDC2022]. Mask wearing has individual benefits, but the CDC emphasized the cooperative aspects of mask wearing as a tool for reducing community spread [@CDC2020]. Indeed, mask wearing posed a social dilemma to individuals, in that it imposed personal costs (e.g., difficulty breathing, disrupted social interaction) for the benefit of the wider community (e.g., "flattening the curve" to protect at-risk individuals). Thus, the evolution of mask wearing in the United States throughout the COVID-19 pandemic allows us to study, on a short timescale within a single population, the emergence of novel injunctive and descriptive social norms and the causal effects of these norms on cooperative behavior.

Recent research has found positive relationships between perceptions of social norms and protective COVID-19 behaviors. In the United States, one cross-sectional study found that perceptions of injunctive norms positively predicted intentions to stay at home to minimize exposure [@Macy2021], and another vignette study found that experimentally-manipulated descriptive norms increased personal mask wearing intentions [@Bokemper2021]. In Germany, a two-wave study found that perceptions of descriptive norms positively predicted future protective behaviors, such as physical distancing [@Rudert2021]. These studies are telling, but since they are experimental, cross-sectional, or only minimally longitudinal, they are unable to distinguish between within-person and between-person change over time [@Hamaker2015], nor do they have the temporal granularity to capture fluctuating changes in norm strength and norm adherence across the entire pandemic. These studies also do not adequately distinguish between the effects of social norms and the effects of other non-social beliefs that are also predicted to shape behavior, such as factual beliefs and personal normative beliefs [@Bicchieri2014].

Here, we use two years of longitudinal data from a representative sample of adults in the United States (18 time points; _n_ = 916) to track the development of descriptive and injunctive mask wearing norms and mask wearing behavior over the course of the COVID-19 pandemic. Between September 2020 and October 2022, we asked participants to report their frequency of mask wearing during in-person interactions, as well as their perceptions of descriptive and injunctive mask wearing norms. We used these data to answer two main research questions. First, how do descriptive and injunctive mask wearing norms emerge and evolve over time in the United States population? Second, how do descriptive and injunctive mask wearing norms causally influence mask wearing behavior over time?

@@

To understand how mask wearing social norms emerged and fluctuated over the course of the COVID-19 pandemic, we first visualized the average descriptive trends of self-reported norm perceptions across the entire study duration. Figure \@ref(fig:plotTimeline) plots self-reported mask wearing behavior and perceptions of descriptive and injunctive mask wearing norms alongside relevant pandemic-related events in the United States, such as CDC public health recommendations and COVID-19 case numbers. These events were obtained from the CDC Museum’s COVID-19 Timeline [@CDC2022] and are included here based on their direct relevance to protective behaviors (e.g., changing CDC guidance on mask wearing, vaccination access, etc.).

(ref:plotTimelineCaption) _Timeline of self-reported mask wearing behavior and perceived social norms in the United States throughout the COVID-19 pandemic._ (a) Points and line ranges indicate mean averages and standard errors for the self-reported mask wearing item. This item was measured across all eighteen time points on a 5-point Likert scale, with higher values indicating increased frequency of personal mask wearing during in-person interactions. (b) Points and line ranges indicate mean averages and standard errors for perceived injunctive mask wearing norms (green) and perceived descriptive mask wearing norms (blue). These items were measured across eleven time points on a 7-point Likert scale, with higher values indicating stronger perceived social norms. (c) Smoothed data for new COVID-19 cases in the United States, displayed on the log scale (data retrieved from Our World in Data; https://ourworldindata.org/). Across all panels, gray dashed lines represent significant pandemic-related events in the United States, such as vaccine approval from the Food and Drug Administration (FDA) and public health recommendations from the Centers for Disease Control and Prevention (CDC).

@@

Two main observations can be made about the emergence and stability of social norms from these visualizations. First, social norms and behavior were tightly coupled over time. Although social norms are measured on fewer occasions than mask wearing behavior, we can see that as mask wearing behavior decreased in the summer of 2021, so too did perceived descriptive and injunctive mask wearing norms. Subsequently, the steep rise in COVID-19 case numbers in the fall of 2021 saw concomitant increases in both mask wearing behavior and perceived social norms, before declining again in 2022. In line with these patterns, multilevel regression models revealed positive correlations between mask wearing behavior and perceived descriptive mask wearing norms (_b_ = `r printnum(fixef(m1.1)["DesNorms"])`, 95% confidence interval [`r printnum(conf1.1["DesNorms","2.5 %"])` `r printnum(conf1.1["DesNorms","97.5 %"])`], _p_ `r printp(summary(m1.1)$coefficients["DesNorms","Pr(>|t|)"])`) and between mask wearing behavior and perceived injunctive mask wearing norms (_b_ = `r printnum(fixef(m1.2)["InjNorms"])`, 95% CI [`r printnum(conf1.2["InjNorms","2.5 %"])` `r printnum(conf1.2["InjNorms","97.5 %"])`], _p_ `r printp(summary(m1.2)$coefficients["InjNorms","Pr(>|t|)"])`) across individuals and time points (Supplementary Figure \@ref(fig:plotCorBehNorm); Supplementary Table \@ref(tab:modelSummaryTable1)).

Second, fluctuations in mask wearing behavior and perceived social norms are in line with recommendations broadcasted by the CDC, the main national public health agency of the United States. We do not have data for the very start of the pandemic in early 2020, but the high levels of mask wearing and strong perceived social norms at the start of our observation window likely emerged after the initial mask wearing recommendation from the CDC in April 2020. Perceived social norms and mask wearing behavior subsequently declined after the CDC rescinded their mask wearing recommendation following widespread vaccine availability in March 2021, and then increased again after the CDC updated their guidelines for indoor mask use in high risk areas in August 2021. Finally, perceived social norms and mask wearing declined again after the CDC eased mask wearing guidelines in March 2022. Impressively, shifts in mask wearing behavior and norms happened on the timescale of weeks. These trends were confirmed by a series of multilevel regression models with change points aligning with changes in CDC mask wearing recommendations (Supplementary Figure \@ref(fig:plotCDCSens); Supplementary Table \@ref(tab:changePointsTable)).

Sample averages can provide informative trends, but they do not allow us to estimate within-person changes in mask wearing behavior and perceived social norms over time. To determine whether within-person changes in social norms temporally preceded within-person changes in mask wearing behavior, we fitted a ten-wave unconstrained random-intercept cross-lagged panel model to the longitudinal data. This structural equation model separately estimated stable trait-like between-person individual differences and within-person fluctuations from trait levels for our main variables (self-reported mask wearing behavior, perceived descriptive mask wearing norms, and perceived injunctive mask wearing norms) and time-variant control variables (factual beliefs and personal normative beliefs). In line with our proposed causal model (Supplementary Figure \@ref(fig:plotDAG)), we also included political orientation as an exogenous time-invariant control. According to established fit statistics, this model fitted the data well (root mean square error of approximation = `r printnum(fitMeasures['rmsea'], digits = 3)`, 95% CI [`r printnum(fitMeasures['rmsea.ci.lower'], digits = 3)` `r printnum(fitMeasures['rmsea.ci.upper'], digits = 3)`]; standardized root mean squared residual = `r printnum(fitMeasures['srmr'], digits = 3)`; comparative fit index = `r printnum(fitMeasures['cfi'], digits = 3)`). Since we are primarily interested in the causal effects of social norms on behavior, in what follows we focus on the results for mask wearing behavior, perceived descriptive norms, and perceived injunctive norms (but see Supplementary Table \@ref(tab:lavaanTable) for full list of estimated autoregressive and cross-lagged effects).

@@

### Self-reported mask wearing behavior

>>>>

---
title             : "Descriptive, not injunctive, social norms caused increases in mask wearing during the COVID-19 pandemic"
shorttitle        : "Norms and mask wearing"

author: 
  - name          : "Samantha L. Heiman"
    affiliation   : "*,1"
    #corresponding : yes
    #address       : "1101 E 10th St, Bloomington, IN 47405, United States"
    #email         : "slheiman@iu.edu"
  - name          : "Scott Claessens"
    affiliation   : "*,2"
  - name          : "Jessica D. Ayers"
    affiliation   : "3"
  - name          : "Diego Guevara Beltran"
    affiliation   : "4"
  - name          : "Andrew Van Horn"
    affiliation   : "5,6"
  - name          : "Edward R. Hirt"
    affiliation   : "1"
  - name          : "Athena Aktipis"
    affiliation   : "†,4"
  - name          : "Peter M. Todd"
    affiliation   : "†,1,7"

affiliation:
  - id            : "1"
    institution   : "\\footnotesize Department of Psychological and Brain Sciences, Indiana University Bloomington, United States"
  - id            : "2"
    institution   : "\\footnotesize School of Psychology, University of Auckland, New Zealand"
  - id            : "3"
    institution   : "\\footnotesize Department of Psychological Science, Boise State University, United States"
  - id            : "4"
    institution   : "\\footnotesize Department of Psychology, Arizona State University, United States"
  - id            : "5"
    institution   : "\\footnotesize Department of Physics, Case Western Reserve University, United States"
  - id            : "6"
    institution   : "\\footnotesize Department of Art History, Case Western Reserve University, United States"  
  - id            : "7"
    institution   : "\\footnotesize Cognitive Science Program, Indiana University Bloomington, United States"

abstract: |
 Human sociality is governed by two types of social norms: injunctive social norms, which prescribe what people _ought_ to do, and descriptive social norms, which reflect what people _actually_ do. These norms enable people to cooperate in the face of group-wide challenges. While previous experimental work has shown that people’s behavior is influenced by social norms, several open questions remain about the natural emergence of injunctive and descriptive social norms within populations and their causal influences on cooperative behavior over time. To understand how cooperative behavior emerges and is shaped by changing social norms in a non-experimental setting, we studied mask wearing during the COVID-19 pandemic. Mask wearing has individual benefits, but it is also a cooperative behavior that provides collective benefits of reduced disease transmission in the community. Leveraging two years of longitudinal data from a representative sample of adults in the United States (18 time points; _n_ = `r nrow(tar_read(d))`), we tracked people’s reported mask wearing and their perceived injunctive and descriptive mask wearing norms as the pandemic unfolded. Longitudinal trends of norm perceptions and self-reported mask wearing suggested that norms and behavior were tightly coupled and both changed quickly in response to recommendations from public health authorities. In addition, a random-intercept cross-lagged panel model revealed that perceived descriptive norms, but not perceived injunctive norms, caused future within-person increases in individuals' mask wearing. These findings show that, during uncertain times, cooperative behavior is driven by what others are actually doing, rather than what others think ought to be done.

note              : "\\footnotesize \\raggedright \\mbox{* indicates shared first authorship, † indicates shared senior authorship} \\par Correspondence concerning this article should be addressed to Samantha L. Heiman, 1101 E 10th St, Bloomington, IN 47405, United States. E-mail: slheiman@iu.edu \\par This study was funded by the Interdisciplinary Cooperation Initiative, ASU President's Office, the Cooperation Science Network, the Institute for Mental Health Research, the University of New Mexico, the Indiana University College of Arts & Sciences, the Rutgers University Center for Human Evolutionary Studies, the Charles Koch Foundation, and the John Templeton Foundation. \\par This working paper has not yet been peer-reviewed."
keywords          : "social norms; descriptive norms; injunctive norms; longitudinal; COVID-19; mask wearing; cooperation"
wordcount         : "5276 words"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man, donotrepeattitle"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
appendix          : "appendix.Rmd"
csl               : pnas.csl
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
tar_load(c(fileData, fileOwid, d, owid,
           constructValDes, constructValInj,
           m1.1, m1.2, m2.1, m2.2, m2.3,
           conf1.1, conf1.2, conf2.1, conf2.2, conf2.3,
           pars2.1, pars2.2, pars2.3, riclpm, fitMeasures,
           plot1, plot2, plot3, plot4, plot5, plot6, plot7,
           itemTable, changePointsTable, lavaanTable))
```

# \normalfont Significance Statement

Social norms have been identified as important drivers of human cooperation, but the emergence of social norms within populations and their subsequent effects on cooperative behavior are not well understood. Here, we use mask wearing during the COVID-19 pandemic as a real-world setting in which to study the emergence of social norms. Over two years in the United States, we find that social norms and mask wearing are tightly coupled and change quickly in response to public health recommendations. Moreover, longitudinal modeling suggests that mask wearing is causally preceded by perceptions of descriptive norms (i.e. what people _are_ doing) but not injunctive norms (i.e. what people _ought_ to be doing).

\begin{center}Descriptive, not injunctive, social norms caused increases in mask wearing during the COVID-19 pandemic\end{center}

Social norms are a key aspect of human sociality [@Bicchieri2009; @Cialdini1991; @Fehr2018a; @VanKleef2019]. Broadly, social norms are defined as commonly known behavioral guidelines enforced by groups of people [@Legros2020]. Since Asch's early studies of normative conformity [@Asch1956], evidence has accrued that humans, unlike other species, are highly attuned to social norms [@Tomasello2014; @Riedl2012]. From a young age, children begin to adhere to and enforce group-wide social norms [@Jensen2014; @Rakoczy2013; @Schmidt2012]. Both children and adults rely on normative emotions, such as shame and guilt, to determine when they or others have violated social norms [@Vaish2011; @Schaumberg2022]. People also readily punish third-party norm violators [@Fehr2004]. Emerging from this uniquely normative psychology, social norms specify the rules and practices that govern human social life, forming the "grammar of society" [@Bicchieri2006].

By coordinating the behavior of many individuals, social norms enable human groups to cooperate in the face of group-wide challenges and threats, such as inter-group conflict, resource scarcity, high population density, natural disasters, and infectious diseases [@Gelfand2011; @Henrich2015; @Roos2015]. In the Maasai people of East Africa, for example, social norms governing transfers of livestock enable individuals to collectively manage the risks associated with nomadic pastoralism (e.g., famine, droughts) [@Aktipis2011]. In countries with high population densities, social norms provide the rules and regularities that allow large groups of people to effectively align their behavior and avoid chaos and conflict [@Gelfand2011]. Social norms are thus hypothesized to have played a key role in the evolution of large-scale cooperation in humans [@Henrich2015; @Chudek2011].

Previous research has distinguished between two primary types of social norms: injunctive norms and descriptive norms [@Bicchieri2009; @Cialdini1991; @Deutsch1955]. Injunctive social norms indicate what others in the group tend to approve or disapprove of and often involve social sanctions if violated. By contrast, descriptive norms simply describe what most people are doing in a given situation. Though these two kinds of social norms tend to align, they can also be in conflict with one another. For example, there may be an injunctive norm that cleaning up litter at a picnic site is the right thing to do: one _ought_ to behave this way. However, if an individual observes that most people are leaving their litter behind at the site, the descriptive norm is to not clean up. It is thus possible for injunctive and descriptive norms to differentially affect behavior [@Schultz2007].

Despite decades of research on the causes and consequences of injunctive and descriptive norms [@Bicchieri2009; @Cialdini1991; @Cialdini2007; @Cialdini2021; @Schultz2007], open questions remain regarding the emergence and causal influence of social norms [@Fehr2018a; @Legros2020; @VanKleef2019]. First, how do injunctive and descriptive norms emerge over time within a population? Second, how do evolving injunctive and descriptive norms causally influence behavior over time?

Research in cultural evolution and behavioral economics has investigated how social norms emerge in a population over time. In the long term, cultural evolutionary models show that injunctive social norms can be vertically transmitted across generations by imitation or teaching, or horizontally diffused from neighboring human populations [@Boyd1985; @Henrich2015]. For example, cultural phylogenetic studies have revealed patterns of vertical cultural inheritance across societies for a variety of injunctive social norms, such as norms governing land ownership [@Kushnick2014] and post-marital residence [@Jordan2009]. However, less is known about how social norms arise endogenously within populations in the short term. Recent experimental work in behavioral economics suggests that social norms of public good provisioning develop in tandem with cooperative behavior through repeated interactions [@Titlestad2019] and require peer enforcement to become stable [@Fehr2018b]. But it remains unclear whether these findings generalize beyond the laboratory to real human populations.

With regards to normative influences on behavior, there is a wealth of cross-sectional evidence demonstrating the impact of social norms on behavior. For example, field experiments in the United Kingdom and United States have demonstrated the positive effects of descriptive norms on a variety of cooperative behaviors, including recycling [@Nigbur2010], paying taxes [@Larkin2018], and sustainably reusing towels in hotels [@Goldstein2008]. Evidence also suggests that any potentially deleterious effects of descriptive social norms (e.g., choosing to litter at a picnic site that already contains visible signs of littering) can be counteracted by instead focusing individuals’ attention on opposing injunctive norms (e.g., seeing that the litter is swept into piles, showcasing that it should be cleaned up) [@Cialdini1991; @Schultz2007].

However, these cross-sectional studies have two main limitations. First, studies have not adequately controlled for other potential non-social influences on behavior, such as perceptions of the effectiveness of the behavior and personal beliefs that one should behave in a certain way. Previous research has labeled these non-social beliefs as "factual beliefs" and "personal normative beliefs" respectively and emphasized that they should be distinguished from injunctive and descriptive social norms [@Bicchieri2014]. For example, willingness to sustainably reuse towels might be driven by perceptions that towel reuse has a positive impact on the environment and/or personal beliefs that towel reuse is the right thing to do, rather than by social norms. Second, cross-sectional studies have tended to follow experimental designs in which perceptions of social norms are manipulated by the researchers at a single time point. But social norms are not static: they change dynamically over time through processes of deliberation and interaction [@Smith2015]. An alternative way to assess causality between social norms and cooperative behavior, while retaining ecological validity, is to follow these variables over time amidst a real, unfolding social dilemma. In response to novel social dilemmas, social norms and cooperative behaviors emerge as people discover the benefits and costs of different behaviors [@Ostrom1990], enabling researchers to take multiple snapshots over time and establish temporal patterns. However, despite the promise of this approach, previous research has not studied the longitudinal interplay between social norms and cooperative behavior.

To understand how novel injunctive and descriptive social norms emerge over time and shape cooperative behavior in a non-experimental setting, in the current work we focus on mask wearing during the COVID-19 pandemic. Before the pandemic, mask wearing was not a common behavior in the United States. In April 2020, one month after the World Health Organization declared COVID-19 a global pandemic, mask wearing was officially recommended by the Centers for Disease Control and Prevention (CDC) as a protective behavior that people should adopt to minimize the spread of the disease [@CDC2022]. Mask wearing has individual benefits, but the CDC also emphasized the collective benefits of the behavior in reducing the spread of the disease throughout the community [@CDC2020]. Indeed, mask wearing posed a social dilemma to many individuals, in that it imposed personal costs (e.g., difficulty breathing, disrupted social interaction) for the benefit of the wider community (e.g., "flattening the curve" to protect at-risk individuals). Thus, the development of mask wearing in the United States during the COVID-19 pandemic allows us to study the emergence of novel injunctive and descriptive social norms and their causal effects on cooperative behavior over a short timescale within a single population.

Recent research has found positive relationships between individuals' perceptions of social norms and their protective COVID-19 behaviors. In the United States, one cross-sectional study found that perceptions of injunctive norms positively predicted intentions to stay at home to minimize exposure [@Macy2021], and another vignette study found that experimentally-manipulated descriptive norms increased personal mask wearing intentions [@Bokemper2021]. In Germany, a two-wave study found that perceptions of descriptive norms positively predicted future protective behaviors, such as physical distancing [@Rudert2021]. These studies are telling, but since they are experimental, cross-sectional, or only minimally longitudinal, they are unable to distinguish between within-person and between-person change over time [@Hamaker2015], nor do they have the temporal granularity to capture fluctuating changes in norm strength and norm adherence across the pandemic. Furthermore, these studies do not distinguish between the effects of social norms and the effects of other non-social beliefs that are predicted to shape behavior, such as factual beliefs and personal normative beliefs [@Bicchieri2014]. Several of the studies also do not control for political ideology, which is important to account for since mask wearing was highly politicized in the United States during the COVID-19 pandemic [@BaxterKing2022].

Here, we use two years of longitudinal data from a representative sample of adults in the United States (18 time points; _n_ = 916) to track the development of descriptive and injunctive mask wearing norms and mask wearing behavior over the course of the COVID-19 pandemic. Between September 2020 and October 2022, we asked participants to report their frequency of mask wearing during in-person interactions, as well as their perceptions of descriptive and injunctive mask wearing norms. Guided by a causal model of social norms and mask wearing (Supplementary Figure \@ref(fig:plotDAG)), we also asked participants about their factual beliefs (i.e., whether they believe mask wearing is effective), personal normative beliefs (i.e., whether they believed mask wearing is the right thing to do), and political ideology, and controlled for these factors. We used our longitudinal data to answer our two main research questions in a specific real-world context. First, how do descriptive and injunctive mask wearing norms emerge and develop over time in the United States population during the pandemic? Second, how do descriptive and injunctive mask wearing norms causally influence mask wearing over time?

@@

To understand how mask wearing social norms emerged and fluctuated over the course of the COVID-19 pandemic, we first visualized the average descriptive trends of self-reported norm perceptions across the entire study duration. Figure \@ref(fig:plotTimeline) plots self-reported mask wearing and perceptions of descriptive and injunctive mask wearing norms alongside relevant pandemic-related events in the United States, such as CDC public health recommendations and COVID-19 case numbers. These events were obtained from the CDC Museum’s COVID-19 Timeline [@CDC2022].

(ref:plotTimelineCaption) _Timeline of self-reported mask wearing and perceived social norms in the United States during the COVID-19 pandemic._ (a) Points and line ranges indicate means ± two standard errors for the self-reported mask wearing item. This item was measured across all eighteen time points on a 5-point Likert scale, with higher values indicating increased frequency of personal mask wearing during in-person interactions. (b) Points and line ranges indicate means ± two standard errors for perceived injunctive mask wearing norms (green) and perceived descriptive mask wearing norms (blue). These items were measured across eleven time points on a 7-point Likert scale, with higher values indicating stronger perceived social norms. (c) Smoothed data for daily new COVID-19 cases in the United States, displayed on the log scale (data retrieved from Our World in Data; https://ourworldindata.org/). Across all panels, gray dashed lines represent significant pandemic-related events in the United States, such as vaccine approval from the Food and Drug Administration (FDA) and public health recommendations from the Centers for Disease Control and Prevention (CDC).

@@

Two main observations can be made about the emergence and stability of social norms from these visualizations. First, social norms and behavior were tightly coupled over time. Although social norms are measured on fewer occasions than mask wearing, we can see that as mask wearing decreased in the summer of 2021, so too did perceived descriptive and injunctive mask wearing norms. Subsequently, the steep rise in COVID-19 case numbers in the fall of 2021 saw concomitant increases in both mask wearing and perceived social norms, before declining again in 2022. In line with these patterns, multilevel regression models revealed positive correlations between mask wearing and perceived descriptive mask wearing norms (_b_ = `r printnum(fixef(m1.1)["DesNorms"])`, 95% confidence interval [`r printnum(conf1.1["DesNorms","2.5 %"])` `r printnum(conf1.1["DesNorms","97.5 %"])`], _p_ `r printp(summary(m1.1)$coefficients["DesNorms","Pr(>|t|)"])`) and between mask wearing and perceived injunctive mask wearing norms (_b_ = `r printnum(fixef(m1.2)["InjNorms"])`, 95% CI [`r printnum(conf1.2["InjNorms","2.5 %"])` `r printnum(conf1.2["InjNorms","97.5 %"])`], _p_ `r printp(summary(m1.2)$coefficients["InjNorms","Pr(>|t|)"])`) across individuals and time points (Supplementary Figure \@ref(fig:plotCorBehNorm); Supplementary Table \@ref(tab:modelSummaryTable1)).

Second, fluctuations in mask wearing and perceived social norms are in line with recommendations broadcasted by the CDC, the main national public health agency of the United States. We do not have data for the very start of the pandemic in early 2020, but the high levels of mask wearing and strong perceived social norms at the start of our observation window likely emerged after the initial mask wearing recommendation from the CDC in April 2020. Perceived social norms and mask wearing subsequently declined after the CDC rescinded their mask wearing recommendation following widespread vaccine availability in March 2021, and then increased again after the CDC updated their guidelines for indoor mask use in high-risk areas in August 2021. Finally, perceived social norms and mask wearing declined again after the CDC eased mask wearing guidelines in March 2022. These trends were confirmed by a series of multilevel regression models with change points aligning with changes in CDC mask wearing recommendations (Supplementary Figure \@ref(fig:plotCDCSens); Supplementary Table \@ref(tab:changePointsTable)).

Sample averages can provide informative trends, but they do not allow us to estimate within-person changes in mask wearing and perceived social norms over time. To determine whether within-person changes in social norms temporally preceded within-person changes in mask wearing, we fitted a ten-wave unconstrained random-intercept cross-lagged panel model to the longitudinal data. This structural equation model separately estimated stable trait-like between-person individual differences and within-person fluctuations from those trait levels for our main variables (self-reported mask wearing, perceived descriptive mask wearing norms, and perceived injunctive mask wearing norms) and time-variant control variables (factual beliefs and personal normative beliefs). In line with our proposed causal model (Supplementary Figure \@ref(fig:plotDAG)), we also included political orientation as an exogenous time-invariant control. According to established fit statistics, this model fitted the data well (root mean square error of approximation = `r printnum(fitMeasures['rmsea'], digits = 3)`, 95% CI [`r printnum(fitMeasures['rmsea.ci.lower'], digits = 3)` `r printnum(fitMeasures['rmsea.ci.upper'], digits = 3)`]; standardized root mean squared residual = `r printnum(fitMeasures['srmr'], digits = 3)`; comparative fit index = `r printnum(fitMeasures['cfi'], digits = 3)`). Since we are primarily interested in the causal effects of social norms on behavior, in what follows we focus on the results for mask wearing, perceived descriptive norms, and perceived injunctive norms (but see Supplementary Table \@ref(tab:lavaanTable) for full list of estimated autoregressive and cross-lagged effects).

@@

### Self-reported mask wearing
